---
title: "Untitled"
output: html_document
---

```{r libraries, echo = FALSE}
library(tidyverse)
library(stringr)
library(gapminder)
library(magrittr)
```

# Lecture 1: Welcome to the tidyverse  

#### read_csv  
Use `readr::read_csv` to bring data from standard comma separated value (.csv) files into R:  
wine_1 <- read_csv("data/wine_1.csv")   # <--- tidy documento in csv  
wine_2 <- read_csv("data/wine_2.csv", skip = 1) # <--- in csv, jump first row  
wine_3 <- read_csv("data/wine_3.csv", n_max = 5) # <--- csv, put a max number of columns  
wine_4 <- read_delim("data/wine_4.txt", delim = "\t") # <--- txt file, delimited by tabs  
wine_5 <- read_csv("data/wine_5.csv", col_types = list("Add more columns" = "-")) # <--- csv, change the name of columns to "-" to be erased  
wine_6 <- read_delim("data/wine_6.txt", delim = "\t",  skip = 1,  col_types = list("C" = "-", "X10"= "-", "R" = "-"),  col_names = TRUE,  n_max = 5)  
wine_7 <- read_excel("data/wine_7.xlsx") # <--- excel   
wine_8 <- read_excel("data/wine_8.xlsx", sheet = "DATA") # <--- excel in a specific spreadsheet  
  
  
#### What is `tidy data`?
A tidy data is one that is satified by these three criteria:  
- each row is a single observation,  
- each variable is a single column, and  
- each value is a single cell (i.e., its row, column position in the data frame is not shared with another value)  
*What is a variable and an observation may depend on your immediate goal.*  
  
Five most common problems with messy datasets, along with their remedies:  
• Column headers are values, not variable names.  
• Multiple variables are stored in one column.  
• Variables are stored in both rows and columns.  
• Multiple types of observational units are stored in the same table. • A single observational unit is stored in multiple tables.  
  
#### Advantages and disadvantages of the tidy data format  
Tidy data is only worthwhile if it makes analysis easier.

#### gather and spread
gather(data, key = xxx, value = yyyy, `t1`:`t9`) #columns t1 to t9
gather(data, key = xxx, value = yyyy, -CountryName) #all columns exept the last one
- key: the name of a new column that will be created, whose values will come from the names of the columns that we want to combine.
- values: the name of a new column that will be created, whose values will come from the values of the columns we want to combine.
- names of the columns that we want to combine.


spread(data, key = xxx, value = yyyy)
- key: the name of the column whose values we would like to use as column names when we spread the data.
- value: the name of the column whose values you would like to spread into separate columns.


#### select: Use select to subset the data on variables or columns.
select(gapminder, year, lifeExp)

#### filter: takes logical expressions and returns the rows for which all are TRUE.
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Rwanda", year > 1979)
filter(gapminder, lifeExp > 80 | year == 2007)
filter(gapminder, country `%in%` c("Rwanda", "Afghanistan"))

#### mutate: to add new variables, or change a column in place
mutate(tidy_unicef, year = as.numeric(sub("U5MR ", "", year)))
gapminder  %>%  mutate(gdp = pop * gdpPercap) 
gapminder  %>%  mutate(gdp = pop * gdpPercap,  pop_thousands = pop / 1000)
gapminder  %>%  mutate(lifeExp = round(lifeExp, 0))

#### %in%: to filter for multiple elements in a column
filter(gapminder, country `%in%` c("Rwanda", "Afghanistan"))

#### str: see the structure of the data
str(gapminder)

#### pipe operator, %>%
The pipe operator takes the thing on the left-hand-side and pipes it into the function call on the right-hand-side – literally, drops it in as the first argument.
gapminder  %>%  filter(country == "Cambodia")  %>%  select(year, lifeExp)



# Lecture 2: Common data manipulations & reading data into R

**Note: if a variable is a factor, we can't manipulate them. So, we need to change these to character vectors so we can manipulate the character values they hold:**
gap <- gapminder %>% 
   mutate(country = as.character(country),
   continent = as.character(continent))
  
#### readr::read_* 
See lecture 1, first part.  
  
#### if_else & case_when


#filter(gapminder, country == "Cambodia")
gapminder %>%  mutate(country = if_else(country == "Cambodia",  "Kingdom of Cambodia", country))
  
gap <- gap  %>%  mutate(country = if_else(country == "Cambodia",  "Kingdom of Cambodia",  country))
  
  
#### create a tribble
```{r create }
fix_me <- tibble::tribble(
  ~province,             ~cities,
  "BC",                  "Victoria",
  "Alberta",             "Edmonton",
  "SK",                  "Regina",
  "British Columbia",    "Vancouver",
  "Alberta",             "Calgary",
  "BC",                  "Kelowna",
  "AB",                  "Lethbridge",
  "SK",                  "Saskatoon"
)
fix_me
```

#### removing rows with NA
df %>% drop_na() # <--- from all the tibble  
df %>% drop_na(y) #<--- from only the y variable  
  
#### factors
explore with: str(gapminder$continent)  
> Factor w/ 5 levels "Africa","Americas",..  
  
levels(gapminder$continent)  
>'Africa'  
'Americas'  
'Asia'  
'Europe'  
'Oceania'  
  
nlevels(gapminder$continent)  
>5
  
class(gapminder$continent)  
>'factor'  
  
####Dropping unused levels  
h_countries <- gapminder %>%   
filter(country %in% c("Egypt", "Haiti", "Romania", "Thailand", "Venezuela"))  
  
h_countries$country %>%  
fct_drop() %>%  
nlevels
  
#### Change order of the levels, principled¶
By default, factor levels are ordered alphabetically.  
Change by Frequency (use: `fct_infreq`):  
```{r}
gapminder$continent %>% 
    fct_infreq() %>% 
    levels()
```
  
Change by reverese frequency (use: `fct_rev`):  
```{r}
gapminder$continent %>% 
  fct_infreq() %>%
  fct_rev() %>% 
  levels()
```
  
#### Order one variable by another (reorder)
```{r}
## order accoring to max life exp instead of median
fct_reorder(gapminder$country, gapminder$lifeExp, min) %>% 
    levels() %>% 
    head()
```


#### Loading/importing data¶
  
#### Cleaning up column names¶
library(janitor)  
clean_names(census)  
  
One of the first things you want to do when you read in data can be to clean these up!  
  
If things are not too bad, and there are not too many, `rename` can be a friend.  
```{r}
# rename column names
gapminder %>% 
    rename(life_exp = lifeExp,
          gdp_percap = gdpPercap) %>% 
    head()
```
  
Cleaning up a lot of column names: Sometimes there are A LOT of column names with non-syntactic names. rename is not the best choice here. Instead we'll use the `clean_names` from the janitor package to transform all column names to syntactic ones. 
```{r}
library(janitor)
#census <- read_csv("data/census_snippet.csv")
#head(census)
gapminder
clean_names(gapminder) %>% head()
```


# Lecture 3 - Strings, dates & times, as well as group_by + summarise for iteration

#### String manipulation
Cheatsheet: https://stringr.tidyverse.org  
  
Compare if the string has some text:  
`str_detect`(fruit, "fruit")  
  
Return only the strings that cointain the text:  
`str_subset`(fruit, "fruit")  
  
Split the words string in a list separating the strings that are separated by " ".  
`str_split`(my_fruit, " ")  
  
and this split separate the string in a fixed number of columns.:
`str_split_fixed`(my_fruit, pattern = " ", n = 2)  
  
iunsplit returns a data frame with the information  
tibble(unsplit = my_fruit) %>%   
separate(`unsplit`, into = c("pre", "post"), sep = " ")  

get length of each string  
`str_length`(my_fruit)  
  
remove first three characters  
`str_sub`(my_fruit, 1, 3)  
  
collapse a character vector into one 
head(fruit) %>% str_c(`collapse` = "-")

replace fruit with vegetable  
str_replace(my_fruit, "fruit", "vegetable"  
  
detect countries that start with "AL":
gapminder %>%  
filter(`str_detect`(country, "^Al")) %>%  
pull(country)  %>%  unique()  %>%  length()
  
detect countries that end with "tan"
gapminder  %>%  filter(str_detect(country, "tan$"))  
  
detect countries that contain ", Dem. Rep." 
gapminder  %>%  filter(str_detect(country, "\\, Dem. Rep."))
  
replace ", Dem. Rep." with " Democratic Republic"
gapminder %>% 
  mutate(country = `str_replace`(country, "\\, Dem. Rep.", " Democratic Republic")) %>%
  filter(country == "Korea Democratic Republic")
  
  
### Summary of string manipulation functions covered so far:

| function | description |
|----------|-------------|
| `str_detect` | Detects elements in a vector that match a pattern, returns a vector of logicals |
| `srt_subset` | Detects and returns elements in a vector that match a pattern |
| `str_split` | Split strings in a vector on a delimiter. Returns a list (used `str_split_fixed` to get a matrix) |
| `separate` | Split character vectors from a data frame on a delimiter which get returned as additional columns in the data frame |
| `str_length` | Counts the number of characters for each element of a character vector, and returns a numeric vector of the counts |
| `str_sub` | Remove substrings based on character position |
| `str_c` | Collapse and/or concatenate elements from a character vector(s) |
| `unite` | Concatenate elements from character vectors from a data frame to create a single column |
| `str_replace` | Replace a pattern in a vector of character vectors with a given string |
| `str_extract` | Extract the actual text of a match from a character vector |
| `str_match` | Use capture groups to extract parts of a complex match from a character vector, returns the match and the capture groups as columns of a matrix |
  
  
#### dates and times using the lubridate package
```{r}
library(lubridate)  #ladi the package lubridate

class(today()) #<--- 'Date'
class(now()) #<--- POSIXct' 'POSIXt'

ymd("2017-01-31")  
mdy("January 31st, 2017")  
dmy("31-Jan-2017")  
ymd(20170131)  
ymd_hms("2017-01-31 20:11:59")  
mdy_hm("01/31/2017 08:01")  
```

If you have a data frame with dates separated by month year and day, you can uso:
dates  %>%  mutate(date = `make_date`(year, month, day))
  
getting components:
```{r}
datetime <- now()
datetime
year(datetime)  
month(datetime)  
mday(datetime)  
yday(datetime)  #day of the year
wday(datetime, label = TRUE, abbr = FALSE)  #day of the week
hour(datetime)
minute(datetime)
second(datetime)

born <- dmy(30121982)
age = today() - born
paste("My age:", as.duration(age))
```

#### Iteration with group_by + summarise¶
##### `summarise`
calculates summaries over rows
  
##### Iteration with `group_by + summarise`
Useful when you want to do something repeatedly to a group of rows  
  
Example:
calculate the average life expectancy for each continent  
gapminder %>%   
    group_by(continent) %>%   #<---also I can use it for multiple columns: group_by(continent, year)  
    summarise(mean_life_exp = mean(lifeExp))  
  
#### Grouped mutate
If you don’t want to collapse the 'n' rows for each group into one row, but you want to keep your groups and compute within them:
```{r}
# calculate life expectancy gained (or lost) relative to 1952
gapminder %>% 
    group_by(continent) %>% 
    mutate(life_exp_gain = lifeExp - first(lifeExp)) %>% 
    head()
```
  
  
  #### A madeup example with the material, by Andres
```{r students}
UBC_students <- tibble::tribble(
  ~Name, ~Argum, ~Stats, ~Wrang,
  'Manuel',   92,  87,  90,
  'Rob',     100, 100, 100,
  'Vic',      94,  90,  78,
  'Andres',   85,  70,  90
)

one_more <- c("Gigi", "Wrang", 94)

UBC_students2 <- UBC_students %>% gather(key = class, value = grade, `Argum`:`Wrang`) %>% 
  rbind(one_more) %>% 
  mutate(random_num = rnorm(1, 0, 1))

UBC_students2
#UBC_students2 %>% spread(key = class, value = grade) select("Name", "Wrang", "Argum", "Stats", "random_num")
```
  
  
# Lecture 4 - Nested data frames and combinging data frames (binds and joins)
  
#### Nested data frames: `nest()`
```{r}
by_country <- gapminder %>% 
group_by(continent, country) %>% 
nest() %>% 
head()
print(by_country)
```
The data column is actually a list of data frames (or tibbles, to be precise). This seems like a crazy idea: we have a data frame with a column that is a list of other data frames!  
  
Difference between a standard grouped data frame and a nested data frame: in a grouped data frame, each row is an observation; in a nested data frame, each row is a group.  
  
Another way to think about a nested dataset is we now have a meta-observation: a row that represents the complete time course for a country, rather than a single point in time.
  
#### List-columns in data frames versus tibbles
1. Create a list column using a function such as `nest` (other possibilities are `summarise` + `list`, or `mutate` + a `map_*` function, as described in [Creating list-columns](https://r4ds.had.co.nz/many-models.html#creating-list-columns))  
2. Create other intermediate list-columns by transforming existing list columns with `map`, `map2` or `pmap`  
3. Simplify the list-column back down to a data frame or atomic vector, often by `unnest`, `mutate` + `map_*` functions that return atomic vectors as opposed to lists.  
  
#####################

test



```{r}
library(tidyverse)
library(gapminder)

gapminder %>% 
  mutate(gapminder, gdp_total = pop * gdpPercap)
str(gapminder)  

```



```{r}
gapminder %>% 
select(-year)

grades_data <- read_csv("grades.csv", skip = 2, col_names = TRUE, )
```





```{r}
gapminder %>% group_by(year) %>%
  summarize(min_life_Exp = min(lifeExp))
```



```{r}
gapminder %>% 
  mutate(gapminder, gdp_total = pop * gdpPercap)

```